---
title: "Assignment 3 - EDA"
author: "Nick Sun"
date: "May 26, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning=FALSE,
                      message=FALSE,
                      out.width="5in",
                      fig.align="center")
library(tidyverse)
library(jsonlite)
library(ggrepel)
source("assignment3_functions.R")

p5 <- c("Big Ten", "Big 12", "ACC", "Pac-12", "SEC")
```

#### College Football Analysis - Introduction

The data I used is kept on a public API and contains everything from game scores, player stats, coaches salaries, talent rankings, to AP rankings (complete documentation can be found on https://api.collegefootballdata.com/api/docs/).
I retrieved my data using the `jsonlite` R package and cleaned it using `dplyr`.
I have also written several functions to hit certain API endpoints and have attached them as a `.R` file which I `source()` in this document.
This API is maintained by Reddit user `/u/BlueSCar` who also serves as moderator for the `/r/CFBAnalysis` subreddit.

#### Play Call Effectiveness between Conferences

In my exploratory investigation, it quickly became apparent that **all** the conferences have roughly the same play-calling style - approximately 50% passing and 50% rushing.
What was more interesting was the effectiveness of different types of plays among the different conferences.

```{r pbpdata}
# p5 <- c("Big Ten", "Big 12", "ACC", "Pac-12", "SEC")
# 
# big10 <- get_conf_pbp("B1G")
# acc <- get_conf_pbp("ACC")
# big12 <- get_conf_pbp("B12")
# pac12 <- get_conf_pbp("PAC")
# sec <- get_conf_pbp("SEC")
# 
# pbpdata <- rbind(big10, acc, big12, pac12, sec)
# pbpdata <- pbpdata[!duplicated(pbpdata),]
# 
# write.csv(pbpdata, "pbpdata.csv")
pbpdata <- read.csv("pbpdata.csv")
```

```{r out.height="3in",out.width="5in", fig.align="center"}
pbpdata %>% 
  filter(offense_conference %in% p5) %>% 
  filter(play_type %in% c("Rush", "Rushing Touchdown", "Pass Reception", "Passing Touchdown")) %>% 
  group_by(offense_conference, play_type) %>% 
  summarise(
    yardage = sum(yards_gained),
    plays = n(),
    ypp = yardage/plays
  ) %>% 
  mutate(
    runorpass = ifelse(play_type %in% c("Pass Reception", "Passing Touchdown"), "Pass", "Rush")
  ) %>% 
  ggplot() + 
  geom_bar(mapping = aes(x = play_type, y = ypp, fill = play_type), stat="identity") + 
  facet_grid(
    cols = vars(offense_conference)
  ) +
  labs(title = "The Majority of College Football Scoring is Explosive",
       subtitle = "Plays which score have over double the average yardage than those which do not",
       y = "Average Yards per Play",
       x = "Types of Plays",
       caption = "Among P5 Conferences in 2018, there is not much difference in average play effectiveness") +
  scale_fill_manual(
    "Types of Plays",
    values=c("#D55E00","#E69F00",
             "#0072B2","#56B4E9"),
    labels = c("Pass", "Passing TD", "Rush", "Rush TD")
  ) +
  theme(
    axis.text.x = element_blank()
  )

```

Surprisingly, there is *little to no* practical difference between any of the conferences.
Also in each conference *scoring plays have more than twice the average yardage*.
I think this speaks to the immaturity of all college defenses to read offenses which results in most of the scoring in college football.

#### Conference competitiveness

Each of the Power 5 conferences claims to be balanced and competitive from top to bottom.
I explored this using average margin of victory for matchups bewteen and within the conferences.

```{r scoredata}
# big10scores <- pythag_points("b1g")
# accscores <- pythag_points("acc")
# b12scores <- pythag_points("b12")
# secscores <- pythag_points("sec")
# pacscores <- pythag_points("pac")
# 
# scoredata <- rbind(big10scores, accscores, b12scores, secscores, pacscores)
# 
# big10scores <- pythag_points("b1g", "2017")
# accscores <- pythag_points("acc", "2017")
# b12scores <- pythag_points("b12", "2017")
# secscores <- pythag_points("sec", "2017")
# pacscores <- pythag_points("pac", "2017")
# 
# scoredata2 <- rbind(big10scores, accscores, b12scores, secscores, pacscores)
# 
# big10scores <- pythag_points("b1g", "2016")
# accscores <- pythag_points("acc", "2016")
# b12scores <- pythag_points("b12", "2016")
# secscores <- pythag_points("sec", "2016")
# pacscores <- pythag_points("pac", "2016")
# 
# scoredata3 <- rbind(big10scores, accscores, b12scores, secscores, pacscores)
# 
# big10scores <- pythag_points("b1g", "2015")
# accscores <- pythag_points("acc", "2015")
# b12scores <- pythag_points("b12", "2015")
# secscores <- pythag_points("sec", "2015")
# pacscores <- pythag_points("pac", "2015")
# 
# scoredata4 <- rbind(big10scores, accscores, b12scores, secscores, pacscores)
# 
# big10scores <- pythag_points("b1g", "2014")
# accscores <- pythag_points("acc", "2014")
# b12scores <- pythag_points("b12", "2014")
# secscores <- pythag_points("sec", "2014")
# pacscores <- pythag_points("pac", "2014")
# 
# scoredata5 <- rbind(big10scores, accscores, b12scores, secscores, pacscores)
# 
# scoredata_final <- rbind(scoredata, scoredata2, scoredata3, scoredata4, scoredata5)
# write.csv(scoredata_final, "scoredata.csv")

scoredata_final <- read.csv("scoredata.csv")
```

```{r marginofvictory, out.height="4in", out.width="4in", fig.align = "center"}
scoredata_final %>% 
  group_by(winning_conference, losing_conference) %>% 
  summarise(
    average_margin = mean(margin_of_victory),
    number_of_games = n()
  ) %>% 
  filter(winning_conference %in% p5 & losing_conference %in% p5) %>% 
  ggplot() +
  geom_tile(mapping = aes(x = winning_conference, y = losing_conference, fill = average_margin)) +
  labs(
    title = "Within-conference Margins of Victory are very similar",
    subtitle = "In out of conference play, the Big 12 wins big.\nThe Pac-12 has to gut them out.",
    x = "Winning Conference",
    y = "Losing Conference",
    caption = "Numbers indicate number of games.\nData is from the 2014-2018 regular seasons."
   ) +
  geom_text(mapping = aes(x = winning_conference, y = losing_conference, label = number_of_games)) +
  scale_fill_continuous(
    "Average\nMargin of Victory", low = "white", high = "red"
  )
```

Intraconference margin of victory is pretty consistent between the conferences, so no conference stands out as being more unbalanced than the others.
In inter-conference games, the Big 12 usually wins big while the Pac-12 usually only edge their opponents.

#### Does talent correspond to a better team?

A popular metric for recruiting is the *team talent composite* where a team with a high composite recruits the best players.
I compared the talent composites of each P5 team with the points they scored in 2018.

```{r teams_by_conference}
# teams_by_conference <- fromJSON("https://api.collegefootballdata.com/teams")
# unique(teams_by_conference$conference)
# 
# teams_by_conference %>% 
#   group_by(conference) %>% 
#   summarise(
#     number_of_teams = n()
#   ) -> number_of_teams_by_conf

# pythagorean_points <- left_join(pythagorean_points, number_of_teams_by_conf)

# pythagorean_points %>% 
#   filter(conference %in% p5) %>% 
#   mutate(
#     points_per_team = `points for`/number_of_teams
#   ) %>% 
#   ggplot() + geom_bar(mapping = aes(x = reorder(conference, desc(points_per_team)), y = points_per_team), stat = "identity")
```

```{r conference_points_by_team}
# big10pbt <- conf_points_by_team("b1g")
# secpbt <- conf_points_by_team("sec")
# accpbt <- conf_points_by_team("acc")
# b12pbt <- conf_points_by_team("b12")
# pacpbt <- conf_points_by_team("pac")
# indpbt <- conf_points_by_team("ind")
# 
# allpbt <- rbind(big10pbt, secpbt, accpbt, b12pbt, pacpbt, indpbt)
# 
# allpbt %>% 
#   filter(total_points > 100) -> allpbt
# 
# allpbt <- arrange(allpbt, desc(total_points))
# allpbt <- allpbt[!duplicated(arrange(allpbt, desc(total_points))$team),]

# write.csv(allpbt, "allpbt.csv")
```

```{r joining talent w points}
# allpbt <- read.csv("allpbt.csv")
# 
# talent <- fromJSON("https://api.collegefootballdata.com/talent?year=2018")
# talent <- talent[,2:3]
# colnames(talent) <- c("team", "talent")
# 
# ptsvstalent <- left_join(allpbt, talent)
# 
# teams_by_conference %>% 
#   filter(conference %in% p5) %>% 
#   select(team, conference) -> teams_by_conference
# 
# left_join(ptsvstalent, teams_by_conference) -> ptsvstalent
# 
# ptsvstalent %>% 
#   filter(!is.na(conference)) -> ptsvstalent
# 
# ptsvstalent$talent <- as.numeric(ptsvstalent$talent)
# 
# write.csv(ptsvstalent, "ptsvstalent.csv")

ptsvstalent <- read.csv("ptsvstalent.csv")
```

```{r talentvspoints, out.height="4in",out.width="6in", fig.align="center"}
ggplot() +
  geom_point(ptsvstalent, mapping = aes(x = talent,
                                        y = total_points,
                                        color = conference),
             size = 2.5) +
  geom_text_repel(ptsvstalent,mapping = aes(x = talent,
                                      y = total_points+7,
                                      label = team),
            size = 1.7) +
  labs(
    title = "For most schools, talent doesn't impact scoring",
    subtitle = "Only Oklahoma, Alabama, Clemson, and Ohio St are in a league apart",
    y = "Total Points Scored in 2018"
  ) +
  theme_minimal() +
  scale_x_continuous("Team Talent Composite Score",
                     breaks = seq(200, 1000, by = 50)) +
  scale_color_manual("Conference",
                    values = c("#E69F00",
                               "#D55E00",
                               "#56B4E9",
                               "#0072B2",
                               "#009E73")) +
  theme(
    legend.position = "top"
  )
```

Talent alone doesn't build championships (as Alabama and company demonstrate, you also probably need money).