---
title: 'R for Data Science: Chapter 3'
author: "Nick Sun"
date: "April 8, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## 3.2.4

```{r, echo=FALSE}
str(mpg)
```

## R Markdown

```{r cars}
ggplot(data =mpg)
```

I SEE NOTHING (of value that is).

```{r}
dim(mpg)
```

THERE ARE 234 ROWS AND 11 COLUMNS.


This is a categorical variable.
**f** stands for front-wheel drive, **r** stands for rear wheel drive, and **4** stands for a four wheel drive car.

```{r, echo=FALSE}
ggplot(data = mpg, aes(x = cyl, y = hwy)) +
  geom_point()
```

```{r, echo=FALSE}
ggplot(data = mpg, aes(x = class, y = drv)) +
  geom_point()
```

This scatterplot really doesn't tell us anything.
Almost all the different classes of cars have some cars that fall into one of the drive categories.

## 3.3.1

```{r, echo=FALSE}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(color = "dodgerblue1")
```

The color aesthetic belongs in the `geom_point()` layer, not inside the `aes()` mapping layer.

```{r, echo=FALSE}
str(mpg)
?mpg
```

Manufacturer, model, cyl, trans, drv, fl, and class are all categorical variables.
The quantitative variables are displ, cty, hwy.
I use the `str()` function which automatically displays the data type of each column in the tibble.

```{r, echo=FALSE}
ggplot(data = mpg, aes(x = displ, y = hwy, color = cty)) +
  geom_point()

ggplot(data = mpg, aes(x = displ, y = hwy, size = cty)) +
  geom_point()

# ggplot(data = mpg, aes(x = displ, y = hwy, shape = cty)) +
#   geom_point()
```

Color mapped to a continuous variable creates a color gradient for each of the points.
Likewise, the size variable creates a size gradient - bigger points correspond to larger values of displ.
A continuous variable cannot be mapped to `shape`.

```{r, echo=FALSE}
ggplot(data = mpg, aes(x = displ, y = displ, color = displ)) +
  geom_point()
```

You can do this, but it creates a pretty useless plot in my opinion.

```{r, echo=FALSE}
ggplot(data = mpg, aes(x = displ, y = hwy, stroke = 1.5)) +
  geom_point(color = "#0038b8", shape = 11, size = 3, fill = "white")
```

The `stroke` aesthetic appears to change the width of the borders of each of the points.

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy, color = displ < 5)) +
  geom_point()
```

It binarizes the continuous variable! Pretty neat. 
