---
title: "Chapter 12.3 Exercises"
author: "Nick Sun"
date: "May 1, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Chapter Notes: Common problems and how to fix em

1. One variable might be spread out across multiple columns
2. One observation might be spread across multiple rows

In `table4a`, we see that the columns 1999 and 2000 are not the names of variables, but rather values of a variable.
Each row represents two observations, not one.

```{r echo=FALSE}
table4a

table4a %>% 
  gather(`1999`, `2000`, key="year", value="cases") -> tidy4a

table4b %>% 
  gather(`1999`, `2000`, key="year", value="population") -> tidy4b 

left_join(tidy4a, tidy4b)
```

Spreading is the opposite of gathering.
You use it when an observations is scattered across multiple rows. 
An example would be `table2`; the `type` column contains the names of columns and the values of those columns are held in the `count` column.

```{r echo=FALSE}
table2

table2 %>% 
  spread(key = `type`, value = `count`)
```

### Question 1

Why are `gather` and `spread` not perfectly symmetrical?
Consider the following problem:

```{r echo=FALSE}
stocks <- tibble(
  year   = c(2015, 2015, 2016, 2016),
  half  = c(   1,    2,     1,    2),
  return = c(1.88, 0.59, 0.92, 0.17)
)

stocks

stocks %>% spread(year, return)

stocks %>% 
  spread(year, return) %>% 
  gather("year", "return", `2015`:`2016`)
```

The table `stocks` originally has an observation spread out over multiple rows.
`Spread` makes it so that `year` is now split into individual columns with `return` as the values in those columns.
Running `gather` on this makes a brand new column `year` that is different from the original data in that it is a character vector instead of a dbl vector.

The `convert` argument in `gather` and `spread` essentially runs another function `type_convert` on the new columns or rows.
This function converts data to appropriate types, for example logicals, integers, etc.
If we rerun our earlier code with `convert=TRUE`, we get the following:

```{r echo=FALSE}
stocks %>% 
  spread(year, return, convert=TRUE) %>% 
  gather("year", "return", `2015`:`2016`, convert=TRUE)
```

The output is largely unchanged except that `year` is an integer column instead of character.

### Question 2: Why does this code fail?

```{r }
# table4a %>% 
#   gather(1999, 2000, key = "year", value = "cases")

table4a %>% 
  gather(`1999`, `2000`, key = "year", value = "cases")
```

The column names have to be within backticks.
Otherwise, R thinks that 1999 and 2000 are just integers.
You don't need backticks if the name of the columns are characters.

### Question 3 

Why does spreading this tibble fail? How could you add a new column to fix this issue?

```{r}
people <- tribble(
  ~name,             ~key,    ~value,
  #-----------------|--------|------
  "Phillip Woods",   "age",       45,
  "Phillip Woods",   "height",   186,
  "Phillip Woods",   "age",       50,
  "Jessica Cordero", "age",       37,
  "Jessica Cordero", "height",   156
)

people

# people %>% 
#   spread(key = key, value = value)
```

We get an error here because each row of output must be unique.
Rows 1 and 3 have the same keys - they both have the age of Philip Woods.
To fix this, we can add a column that has distinct values.

```{r echo=FALSE}
people %>% 
  group_by(name ,key) %>% 
  mutate(rownumber = row_number()) %>% 
  spread(key = key, value = value)
```


### Question 4

Tidy the simple tibble below.
Do you need to spread or gather it?
What are the variables?

The rules for tidy data are as follows:

1. Each variable must have its own column.
2. Each observation must have its own row.
3. Each value must have its own cell.

```{r echo=FALSE}
preg <- tribble(
  ~pregnant, ~male, ~female,
  "yes",     NA,    10,
  "no",      20,    12
)

preg
```

We can gather `male` and `female` into one column called `sex`.
We can also remove the NA observation by setting na.rm equal to TRUE.

```{r echo=FALSE}
preg %>% 
  gather(male, female, key = sex, value = count, na.rm = TRUE)
```

