---
title: "Shiny Tutorial"
author: "Nick Sun"
date: "June 1, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(shiny)
```

# Basic Shiny Architecture

Garrett recommends starting off each app using this same basic template:

```{r cars}
ui <- fluidPage(
  # Input functions - every function needs an input name and a label
  # Example:
  sliderInput(inputID = "num", label = "Choose a number")
  
  # Output functions
  plotOutput("hist")
)

server <- function(input, output){}

shinyApp(ui = ui, server = server)
```
We add elements using fluidPage which contains both inputs and outputs.
For our inputs in our CFB data, we probably want to use a selectInput() to pick a conference and then checkboxGroupInput() to make the team selectors within each conference.

```{r server}
server <- function(input, output) {
  output$hist <- renderPlot({
    hist(rnorm(input$num))
  })
}

```

For the server function, we always save objects to `output$[name of object in fluidPage]`.
We build objects in the server function using `render...` functions, for example `renderPlot({})`.
The curly braces allow us to put multiple lines of R code inside of the render function.

We can access our input values within the server function using `input$[object of input]`.

When sharing your app, you have to either name your app as `app.R` or split the app into two files for the ui and server functions: `ui.R` and `server.R` respectively.

You can deploy apps directly to `shinyapps.io` using the following command:

```{r}
library(rsconnect)
rsconnect::deployApp('path/to/your/app')
```
